---
title: "Wood Thrush Site Fidelity Data Exploration"
author: "Jasmine Fu"
date: "2023-05-27"
output:
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Read in migrant data + Load tidyverse

source('source_script.R')

# Load necessary packages for spatial data

library(sf)
library(tmap)

# Read in net data

nets <-
  read_csv("trbo_nets.csv",
           n_max = 30) %>%
  select(net_number = `Net Number`,
         longitude = Long,
         latitude = Lat) %>%
  
  # Turn latitude and longitude into sf objects
  
  mutate(
    longitude = as.numeric(longitude),
    latitude = as.numeric(latitude)) %>%
  st_as_sf(coords = c("longitude", "latitude"),
           crs = 4326) %>%
  select(net_number,
         gps = geometry)

```
## Introduction

In this analysis, I attempt to find evidence for site fidelity at TREES in Wood Thrushes.

## Subsetting the population to recaptures

I dropped observations that did not have a valid net number. Then, I subset the data to all WOTH individuals that were captured more than 4 times. I replaced the net numbers with the the GPS coordinates of each net. The column `captures` shows the number of times an individual has been captured at the TREES site.

```{r recapture-pts}
# Get the band numbers of recaptured individuals

recaps <-
  migrant %>%
  filter(code %in% c("R", "SD")) %>%
  distinct(band_full)

# Get the GPS points of each observation

recapture_pts <- 
  migrant %>%
  
  # Retain individuals that were captured more than once
  
  semi_join(recaps,
            by = "band_full") %>%
  
  # Drop observations with no valid net number
  
  drop_na(net_number) %>%
  
  # Create 'captures': no. of times an individual has been captured
  
  mutate(captures = n(),
         .by = band_full) %>%
  
  # Keep WOTH individuals that have been captured more than 4 times
  
  filter(
    species_code %in% "WOTH",
    captures > 4) %>%
  
  # Get GPS coordinates from 'net' data
  
  inner_join(nets, by = "net_number") %>%
  
  # Keep the following columns
  
  select(band_full, captures, gps)

recapture_pts
```

The data I will work with in the next steps consists of `r nrow(recapture_pts)` observations from `r recapture_pts %>% distinct(band_full) %>% nrow()` individuals (as seen above).

## Calculating centroids

I took the GPS points for each individual and calculated their center point (i.e. centroid). The map below visualizes the centroid of each individual in yellow.

```{r centroid-map}
centroids <-
  recapture_pts %>%
  summarize(centroid = st_union(gps) %>%
              st_centroid(),
            .by = band_full) %>%
  left_join(recapture_pts %>%
              distinct(band_full, captures),
            by = "band_full")

# A map of the nets and individual's centroids

tmap_mode("view")

# Add the tmap layer for points

tm_basemap("Esri.WorldImagery") +
  tm_shape(nets) +
  tm_dots(size = 0.15, col = "white") +
  tm_text("net_number", size = 0.75) +
  tm_shape(st_as_sf(centroids)) +
  tm_dots(col = "yellow", size = 0.1, alpha = 0.7)
```

## Calculating average distance from centroids

I calculated the distance between the GPS coordinates of the observation and the centroid of the individual. Then I calculated the mean distance from centroid for each individual. The mean distances are plotted in the histogram below.

```{r distances-indv}
# Get the mean distance from centroids for each individual

distances_indv <-
  recapture_pts %>%
  select(band_full, gps) %>%
  
  # Calculate centroid for each individual
  
  group_by(band_full) %>%
  mutate(centroid = st_union(gps) %>%
           st_centroid()) %>%
  ungroup() %>%
  
  # Calculate distance from centroid for each individual
  
  rowwise() %>%
  mutate(distance = st_distance(gps, centroid)) %>%
  ungroup() %>%
  
  # Calculate the mean distance from centroid for each individual
  
  summarize(mean_distance = mean(distance) %>%
              as.numeric(),
            .by = band_full)
  
# Get the mean distance from centroid for all individuals

actual_mean <-
  distances_indv %>%
  summarize(avg = mean(mean_distance)) %>%
  pull(avg)

# Plot distances

ggplot(distances_indv, aes(x = mean_distance)) +
  geom_histogram(binwidth = 10) +
  geom_vline(xintercept = actual_mean,
             color = "red") +
  annotate("text",
           x = actual_mean + 2,
           y = 1.5,
           label = "Mean distance for all individuals",
           angle = 90,
           color = "white") +
  scale_y_continuous(expand = c(0,0),
                     limits = c(0,4)) +
  labs(title = "Mean distance from centroids by individual",
       x = "Mean distance from centroid (m)",
       y = "Number of individuals")

```

The mean distance from the centroid for all individuals is `r actual_mean` meters, and the median is `r distances_indv %>% summarize(median = median(mean_distance)) %>% pull(median)` meters.

## Permutation test
I performed a permutation test to verify if net visits are dependent on individual. If individuals are randomly visiting nets, then permuting (mixing up) the net numbers will yield a similar mean distance to the actual mean.

If individuals do not have a specific preference for a small area, then the actual mean will be significantly smaller than the mean from permuting the net numbers in the data.

In the code below, I permuted the net numbers of the data, then calculated the mean distance from the centroid of each individual. I repeated this process 999 more times to generate 1000 permuted means in total. I plotted these means with the actual mean of the data in a histogram.

```{r permutation-test}

# Set seed for reproducibility
set.seed(54)

# Perform the permutation and calculation a thousand times

permuted_means <-
  
  # Perform the operation below 1000 times
  
  replicate(1000, {
    
    # Filter data for recaptured individuals
    
    migrant %>%
      semi_join(recaps,
                by = "band_full") %>%
      
      # Drop observations that have an invalid net number
      
      drop_na(net_number) %>%
      
      # Retain WOTH individuals with more than 4 captures
      
      filter(species_code %in% "WOTH") %>%
      filter(n() > 4,
             .by = band_full) %>%
      
      # Permute the net numbers of their observations
      
      mutate(net_number = sample(net_number)) %>%
      
      # Get GPS points of nets
      
      inner_join(nets, by = "net_number") %>%
      
      # Calculate the centroid of individuals
      
      group_by(band_full) %>%
      mutate(centroid = st_union(gps) %>%
               st_centroid()) %>%
      ungroup() %>%
      
      # Calculate the distance from centroid
      
      rowwise() %>%
      mutate(distance = st_distance(gps, centroid)) %>%
      ungroup() %>%
      
      # Calculate the mean distance for each individual
      
      summarize(mean_distance = mean(distance),
                .by = band_full) %>%
      
      # Calculate the average mean distance for all individuals
      
      summarize(avg = mean(mean_distance)) %>%
      pull(avg)
  })


# Create a histogram of the average values
ggplot(data.frame(
  avg = unlist(permuted_means)),
  aes(x = avg)) +
  geom_histogram(binwidth = 0.5) +
  geom_vline(xintercept = actual_mean,
             color = "red") +
  annotate("text",
           x = actual_mean + 1,
           y = 50,
           label = "Mean distance for all individuals",
           angle = 90,
           color = "black") +
  scale_y_continuous(expand = c(0,0),
                     breaks = seq(0,80,10),
                     limits = c(0,80)) +
  labs(title = "Permuted means vs. actual mean",
       x = "Permuted mean distance (m)",
       y = "Number of results")
```

As seen above, the actual mean is a lot lower than the means calculated from the permutation test.

### Statistical difference of permuted and actual mean
I performed a right-tailed student's t-test to verify that the permuted means are statistically higher than the actual mean in the data.
```{r t-test}
# Test statistical difference of original and permuted means
t.test(permuted_means,
       
       # Perform a right-tailed t-test
       
       alternative = "greater",
       mu = actual_mean)
```